
---
title: "nfl_data_predictor_viz"
author: "Jeremy"
date: "9/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Dependencies
```{r}
# install.packages('devtools')
library(devtools) 
install_github("AppliedDataSciencePartners/xgboostExplainer")
library(xgboost)
library(caret)
library(OptimalCutpoints)
library(xgboostExplainer)
library(pROC)
library(SHAPforxgboost)
library(randomForest)
library(rpart) 
library(splitstackshape)
library(tidyverse)
library(Metrics)
```

## Creating the Play-by_Play Aggregations with a 5 Week Lag
```{r}
# Load Data
# load("nfl_pbp_2018_2020.rda")
# Extract Games
games <- unique(pbp_2018_2020[, c(2,4,5,7,284,322, 323)])

# Drop games after week 15
pbp_use <- pbp_2018_2020[pbp_2018_2020$week <= 15,]


# Create index for play by play
pbp_index <- rep(NA, nrow(pbp_use))

for(i in 1:nrow(pbp_use)){
  if(pbp_use$season[i] == 2018){
    pbp_index[i] <- pbp_use$week[i]
  } else if (pbp_use$season[i] == 2019){
    pbp_index[i] <- pbp_use$week[i] + 15
  } else if (pbp_use$season[i] == 2020){
    pbp_index[i] <- pbp_use$week[i] + 30
  }
}

pbp_use$index <- pbp_index

games <- unique(pbp_use[, c(2,4,5,7,284,322, 323, 371)])

#### Define Functions for calculations

epa_calculator_off <- function(data, week_lag = 5, db){
  #'
  #' This function calculates the average EPA for each team for 
  #' different play types. 
  #' 
  #' @param data The play-by-play dataset
  #' @param week_lag The number of weeks to use to calculate averages
  #' @param db The games to create stats for 
  #' 
  #' @return A data frame with the calculated values for each of the teams
  #'
  #'
  
  # Create empty columns to store results
  hrun_epa <- hrun_freq <- hpass_epa <- hpass_freq <- rep(NA, nrow(db))
  hgames <- rep(NA, nrow(db))
  
  arun_epa <- arun_freq <- apass_epa <- apass_freq <- rep(NA, nrow(db))
  agames <- rep(NA, nrow(db))
  
  # For each game
  for(i in 1:nrow(db)){
    # Calculate start index
    start_index <- max(1, db$index[i] - week_lag - 1)
    # Calculate end index
    end_index <- db$index[i] - 1
    # Extract plays involving teams of interest falling in window
    temp <- data[which(data$index > start_index & 
                         data$index <= end_index &
                         data$posteam %in% c(db$home_team[i], db$away_team[i])),]
    # If any such plays exist
    if(nrow(temp) >0 ){
      # Calculate total home run epa
      hrun_epa[i] <- sum(temp$epa[which(temp$play_type == "run" & 
                                          temp$posteam == db$home_team[i])], na.rm = T)
      # Calculate total home run plays
      hrun_freq[i] <- nrow(temp[which(temp$play_type == "run" & 
                                        temp$posteam == db$home_team[i]),])
      # Calculate total home pass epa
      hpass_epa[i] <- sum(temp$epa[which(temp$play_type == "pass" & 
                                           temp$posteam == db$home_team[i])], na.rm = T)
      # Calculate total home pass plays
      hpass_freq[i] <- nrow(temp[which(temp$play_type == "pass" & 
                                         temp$posteam == db$home_team[i]),])
      
      # Calculate total home number of games played
      hgames[i] <- length(unique(na.omit(temp$game_id[which(temp$posteam == db$home_team[i])])))
      
      # Calculate total away run epa
      arun_epa[i] <- sum(temp$epa[which(temp$play_type == "run" & 
                                          temp$posteam == db$away_team[i])], na.rm = T)
      # Calculate total away run plays
      arun_freq[i] <- nrow(temp[which(temp$play_type == "run" & 
                                        temp$posteam == db$away_team[i]),])
      # Calculate total away pass epa
      apass_epa[i] <- sum(temp$epa[which(temp$play_type == "pass" & 
                                           temp$posteam == db$away_team[i])], na.rm = T)
      # Calculate total away pass plays
      apass_freq[i] <- nrow(temp[which(temp$play_type == "pass" & 
                                         temp$posteam == db$away_team[i]),])
      
      # Calculate total number of away games played
      agames[i] <- length(unique(na.omit(temp$game_id[which(temp$posteam == db$away_team[i])])))
    }
    
  }
  # Calculate averages per play
  havg_run_epa <- hrun_epa/hrun_freq
  havg_pass_epa <- hpass_epa/hpass_freq
  
  # Calculate play type frequency per play
  hruns_per_game <- hrun_freq/hgames
  hpass_per_game <- hpass_freq/hgames
  
  # Calculate averages per play
  aavg_run_epa <- arun_epa/arun_freq
  aavg_pass_epa <- apass_epa/apass_freq
  
  # Calculate play type frequency per play
  aruns_per_game <- arun_freq/agames
  apass_per_game <- apass_freq/agames
  
  # Join results columns
  res_dat <- cbind.data.frame(hrun_epa, hrun_freq, hpass_epa, hpass_freq, 
                              havg_run_epa, havg_pass_epa,  hgames,
                              hruns_per_game, hpass_per_game,
                              arun_epa, arun_freq, apass_epa, apass_freq, 
                              aavg_run_epa, aavg_pass_epa,  agames,
                              aruns_per_game, apass_per_game)
  
  # Fill missing with 0
  #res_dat[is.na(res_dat)] <- 0
  # Return Results
  return(res_dat)
}



epa_calculator_def <- function(data, week_lag = 5, db){
  #'
  #' This function calculates the average EPA for each team for 
  #' different play types. 
  #' 
  #' @param data The play-by-play dataset
  #' @param week_lag The number of weeks to use to calculate averages
  #' @param db The games to create stats for 
  #' 
  #' @return A data frame with the calculated values for each of the teams
  #'
  #'
  
  # Create empty columns to store results
  hrun_epa <- hrun_freq <- hpass_epa <- hpass_freq <- rep(NA, nrow(db))
  hgames <- rep(NA, nrow(db))
  
  arun_epa <- arun_freq <- apass_epa <- apass_freq <- rep(NA, nrow(db))
  agames <- rep(NA, nrow(db))
  
  # For each game
  for(i in 1:nrow(db)){
    # Calculate start index
    start_index <- max(1, db$index[i] - week_lag - 1)
    # Calculate end index
    end_index <- db$index[i] - 1
    # Extract plays involving teams of interest falling in window
    temp <- data[which(data$index > start_index & 
                         data$index <= end_index &
                         data$defteam %in% c(db$home_team[i], db$away_team[i])),]
    # If any such plays exist
    if(nrow(temp) >0 ){
      # Calculate total home run epa
      hrun_epa[i] <- sum(temp$epa[which(temp$play_type == "run" & 
                                          temp$defteam == db$home_team[i])], na.rm = T)
      # Calculate total home run plays
      hrun_freq[i] <- nrow(temp[which(temp$play_type == "run" & 
                                        temp$defteam == db$home_team[i]),])
      # Calculate total home pass epa
      hpass_epa[i] <- sum(temp$epa[which(temp$play_type == "pass" & 
                                           temp$defteam == db$home_team[i])], na.rm = T)
      # Calculate total home pass plays
      hpass_freq[i] <- nrow(temp[which(temp$play_type == "pass" & 
                                         temp$defteam == db$home_team[i]),])
      
      # Calculate total home number of games played
      hgames[i] <- length(unique(na.omit(temp$game_id[which(temp$defteam == db$home_team[i])])))
      
      # Calculate total away run epa
      arun_epa[i] <- sum(temp$epa[which(temp$play_type == "run" & 
                                          temp$defteam == db$away_team[i])], na.rm = T)
      # Calculate total away run plays
      arun_freq[i] <- nrow(temp[which(temp$play_type == "run" & 
                                        temp$defteam == db$away_team[i]),])
      # Calculate total away pass epa
      apass_epa[i] <- sum(temp$epa[which(temp$play_type == "pass" & 
                                           temp$defteam == db$away_team[i])], na.rm = T)
      # Calculate total away pass plays
      apass_freq[i] <- nrow(temp[which(temp$play_type == "pass" & 
                                         temp$defteam == db$away_team[i]),])
      
      # Calculate total number of away games played
      agames[i] <- length(unique(na.omit(temp$game_id[which(temp$defteam == db$away_team[i])])))
    }
    
  }
  # Calculate averages per play
  havg_run_epa <- hrun_epa/hrun_freq
  havg_pass_epa <- hpass_epa/hpass_freq
  
  # Calculate play type frequency per play
  hruns_per_game <- hrun_freq/hgames
  hpass_per_game <- hpass_freq/hgames
  
  # Calculate averages per play
  aavg_run_epa <- arun_epa/arun_freq
  aavg_pass_epa <- apass_epa/apass_freq
  
  # Calculate play type frequency per play
  aruns_per_game <- arun_freq/agames
  apass_per_game <- apass_freq/agames
  
  # Join results columns
  res_dat <- cbind.data.frame(hrun_epa, hrun_freq, hpass_epa, hpass_freq, 
                              havg_run_epa, havg_pass_epa,  hgames,
                              hruns_per_game, hpass_per_game,
                              arun_epa, arun_freq, apass_epa, apass_freq, 
                              aavg_run_epa, aavg_pass_epa,  agames,
                              aruns_per_game, apass_per_game)
  # Change names to indicate defense
  names(res_dat) <- paste(names(res_dat), "_def")
  # Fill missing with 0
  #res_dat[is.na(res_dat)] <- 0
  # Return Results
  return(res_dat)
}



### Calculate stats

# Calculate offensive stats
off_stats <- epa_calculator_off(data = pbp_use, week_lag = 5, db = games)

# Calculate defensive stats
def_stats <- epa_calculator_def(data = pbp_use, week_lag = 5, db = games)

mod_dat <- cbind.data.frame(games, off_stats, def_stats)
```

## Wrangling mod_dat to Add Response Variables and Handle Missing Values
```{r}

# Create columns for a winner flag and the point differential
winner <- point_dif <- rep(NA, nrow(mod_dat))

# Loop over each row to determine if the home team won and calculate the point differential
for (i in 1:nrow(mod_dat)) {
  if (mod_dat$home_score[i] > mod_dat$away_score[i]) {
    winner[i] <- 1
    point_dif[i] <- mod_dat$home_score[i] - mod_dat$away_score[i]
  }
  else if (mod_dat$home_score[i] < mod_dat$away_score[i]) {
    winner[i] <- 0
    point_dif[i] <- mod_dat$home_score[i] - mod_dat$away_score[i]
  }
}

# Assign the winner and point_dif vectors to columns in the mod_dat df
mod_dat$winner <- winner
mod_dat$point_dif <- point_dif

# Drop rows with ties
mod_dat <- mod_dat[!is.na(mod_dat$winner), ]

# Fix variable data types
 mod_dat <- mod_dat %>%
   mutate(season = as.factor(season),
          winner = as.factor(winner))
 
 # Eliminate games with index 1 and 2, where there is no data available
 mod_dat2 <- mod_dat[mod_dat$index >= 3, ]
 
 # Scale the data
 mod_dat2 <- cbind( mod_dat2[, 1:8], scale(mod_dat2[, 9:44]), mod_dat2[, 45:46])
```

## Increasing the Sample Size
```{r}
names(mod_dat2)
# Extract game level stats
game_stats <- mod_dat2[,c(1:8, 45,46)]
# Extract home stats
hstats <- mod_dat2[,c(2,9:17, 27:35)]
# Extract away stats
astats <- mod_dat2[,c(3, 18:26,36:44)]

# Fix names
names(hstats) <- names(astats) <- c("team", "run_epa", "run_freq", "pass_epa", "pass_freq",         
                   "avg_run_epa", "avg_pass_epa", "games", "runs_per_game",     
                   "pass_per_game", "run_epa_def", "run_freq_def",     
                   "pass_epa_def", "pass_freq_def", "avg_run_epa_def",
                   "avg_pass_epa_def", 
                   "games_def", "runs_per_game_def", "pass_per_game_def")
# Create copy of home stats
hstats_1 <- hstats
# Append opp to names for copy of home stats
names(hstats_1) <- paste(names(hstats), "_opp", sep = "")

# Create copy of away stats
astats_1 <- astats
# Append opp to names for copy of away stats
names(astats_1) <- paste(names(astats), "_opp", sep = "")

# Join original home and copy of away
data_1 <- cbind.data.frame(hstats, astats_1)
# Join original away and copy of home
data_2 <- cbind.data.frame(astats, hstats_1)

# Create location columns
data_1$location <- "home"
data_2$location <- "away"

# Create winner columns
data_1$winner <- as.numeric(as.character(mod_dat2$winner))
data_2$winner <- abs(1 - as.numeric(as.character(mod_dat2$winner)))

# Create point differential columns
data_1$point_diff <- mod_dat2$point_dif
data_2$point_diff <- -mod_dat2$point_dif

# Rejoin data
use_data <- rbind.data.frame(cbind.data.frame(game_stats, data_1),
                             cbind.data.frame(game_stats, data_2))

# Add relevant interaction terms
use_data$run_int <- use_data$avg_run_epa * use_data$avg_run_epa_def_opp
use_data$pass_int <- use_data$avg_pass_epa * use_data$avg_pass_epa_def_opp
use_data$def_run_int <- use_data$avg_run_epa_def * use_data$avg_run_epa_opp
use_data$def_pass_int <- use_data$avg_pass_epa_def * use_data$avg_pass_epa_opp

# Check stats are correctly linked
use_data[use_data$game_id == "2019_08_TB_TEN",]
```

## Inspect the Response Variable
```{r}
summary(use_data$point_diff)
```

```{r}
plot_data <- use_data[, c(5, 11:51)]
  
g_1 <- ggplot(plot_data, aes(x = avg_run_epa)) +
   geom_density(alpha = 0.4) +
   theme(panel.background = element_blank(),
         panel.border = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) +
   labs(title = 'Distribution of Average Run EPA',
        x = 'Average Run EPA',
        y = 'Frequency')
 g_1
 
 g_2 <- ggplot(plot_data, aes(x = avg_pass_epa)) +
   geom_density(alpha = 0.4) +
   theme(panel.background = element_blank(),
         panel.border = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) +
   labs(title = 'Distribution of Average Pass EPA',
        x = 'Average Pass EPA',
        y = 'Frequency')
 g_2

 g_3 <- ggplot(plot_data, aes(x = point_diff)) +
   geom_density(alpha = 0.4) +
   theme(panel.background = element_blank(),
         panel.border = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) +
   labs(title = 'Distribution of Point Differentials',
        x = 'Point Differential',
        y = 'Frequency')
 g_3

```

## Segment the Dataset
```{r}
# In a new session, reload the pre-saved mod_dat2 dataset
# load('game_level_stats3.rda')
#train_dat <- use_data[use_data$season %in% c('2018', '2019'), ]
 
#test_dat <- use_data[use_data$season == '2020', ]
train_dat <- use_data[use_data$index < 36, ]
 
test_dat <- use_data[use_data$index >= 36, ]

nrow(train_dat)
nrow(test_dat)
```


## XGBoost Prep
```{r}
dtrain <- xgb.DMatrix(data = as.matrix(train_dat[, c(12:29, 31:48, 52:55)]), label = as.numeric(train_dat$point_diff))

dtest <- xgb.DMatrix(data = as.matrix(test_dat[, c(12:29, 31:48, 52:55)]), label = as.numeric(test_dat$point_diff))
```

## XGBoost: Initial Model
```{r}
set.seed(111111)
bst_1 <- xgboost(data = dtrain,
               nrounds = 100,
               verbose = 1,
                print_every_n = 20)

boost_preds <- predict(bst_1, dtest)

pred_dat <- cbind.data.frame(boost_preds, test_dat$point_diff)
names(pred_dat) <- c('prediction', 'actual')
pred_dat$error <- pred_dat$actual - pred_dat$prediction

rmse(boost_preds, test_dat$point_diff)

g_6 <- ggplot(pred_dat, aes(x = error)) +
   geom_density(alpha = 0.4, fill = 'blue') +
   theme(panel.background = element_blank(),
         panel.border = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) +
   labs(title = 'Distribution of Error Terms',
        x = 'Value of Error Term',
        y = 'Frequency')
g_6
```

## XGBoost: Tuning the Number of Iterations

```{r}
set.seed(111111)
bst_mod_1 <- xgb.cv(data = dtrain,
              
              nfold = 5,
               
               eta = 0.005,
              
               nrounds = 1000, 
               early_stopping_rounds = 1000, 
               
               verbose = 1, 
               nthread = 6, 
               print_every_n = 20)

g_6 <- ggplot(bst_mod_1$evaluation_log, aes(x = iter, y = test_rmse_mean)) +
  geom_point(alpha = 0.5, color = "blue") + 
  geom_smooth() + 
  theme_bw() + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank()) + 
  labs(x = "Number of Trees", title = "RMSE v Number of Trees",
       y = "RMSE")
g_6
```

## XGBoost: Tuning the Max Depth and Min Child Parameters
```{r}
max_depth_vals <- c(3, 5, 7, 10, 15) 
min_child_weight <- c(1,3,5,7, 10, 15)


cv_params <- expand.grid(max_depth_vals, min_child_weight)
names(cv_params) <- c("max_depth", "min_child_weight")

rmse_vec <- rep(NA, nrow(cv_params)) 

for(i in 1:nrow(cv_params)){
  set.seed(111111)
  bst_tune <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.005, 
              max.depth = cv_params$max_depth[i], 
              min_child_weight = cv_params$min_child_weight[i], 
             
               
              nrounds = 1000, 
              early_stopping_rounds = 270, 
               
              verbose = 1, 
              nthread = 6, 
              print_every_n = 20)
  
  rmse_vec[i] <- bst_tune$evaluation_log$test_rmse_mean[bst_tune$best_ntreelimit]
  
}
```
```{r}
res_db <- cbind.data.frame(cv_params, rmse_vec)
names(res_db)[3] <- c("test_rmse") 
res_db$max_depth <- as.factor(res_db$max_depth)
res_db$min_child_weight <- as.factor(res_db$min_child_weight)

g_7 <- ggplot(res_db, aes(y = max_depth, x = min_child_weight, fill = test_rmse)) + 
  geom_tile() + 
  theme_bw() +
  scale_fill_gradient2(low = "blue",
    mid = "white", 
    high = "red", 
    midpoint =mean(res_db$test_rmse),
    space = "Lab", 
    na.value ="grey",
    guide = "colourbar", 
    aesthetics = "fill") + 
  labs(x = "Minimum Child Weight", y = "Max Depth", fill = "RMSE")
g_7
```

Minimum Child Weight of 10 and Max Depth of 10 result in the lowest test RMSE.

## XGBoost: Tuning the Gamma Parameter
```{r}
gamma_vals <- c(0, 0.05, 0.1, 0.15, 0.2)

rmse_vec <- rep(NA, length(gamma_vals)) 

set.seed(111111)
for(i in 1:length(gamma_vals)){
  bst_tune <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.005, 
              max.depth = 10, 
              min_child_weight = 10, 
              
              gamma = gamma_vals[i],
             
              nrounds = 1000, 
              early_stopping_rounds = 270, 
               
              verbose = 1, 
              nthread = 6, 
              print_every_n = 20)
  
  rmse_vec[i] <- bst_tune$evaluation_log$test_rmse_mean[bst_tune$best_ntreelimit]
  
}
```
```{r}
cbind(gamma_vals, rmse_vec)
```

Gamma of 0 results in the lowest test RMSE.

## XGBoost: Tuning the Subsample and Column Sample Parameters
```{r}
subsample <- c(0.6, 0.7, 0.8, 0.9, 1) 
colsample_by_tree <- c(0.6, 0.7, 0.8, 0.9, 1)

cv_params <- expand.grid(subsample, colsample_by_tree)
names(cv_params) <- c("subsample", "colsample_by_tree")

rmse_vec <- rep(NA, nrow(cv_params)) 

for(i in 1:nrow(cv_params)){
  set.seed(111111)
  bst_tune <- xgb.cv(data = dtrain, 
              
              nfold = 5,
               
              eta = 0.005, 
              max.depth = 10, 
              min_child_weight = 10,
              
              gamma = 0, 
              subsample = cv_params$subsample[i],
              colsample_bytree = cv_params$colsample_by_tree[i],
               
              nrounds = 1000, 
              early_stopping_rounds = 270, 
               
              verbose = 1, 
              nthread = 6, 
              print_every_n = 20)

  rmse_vec[i] <- bst_tune$evaluation_log$test_rmse_mean[bst_tune$best_ntreelimit]
  
}
```
```{r}
res_db <- cbind.data.frame(cv_params, rmse_vec)
names(res_db)[3] <- c("test_rmse") 
res_db$subsample <- as.factor(res_db$subsample)
res_db$colsample_by_tree <- as.factor(res_db$colsample_by_tree)

g_8 <- ggplot(res_db, aes(y = subsample, x = colsample_by_tree, fill = test_rmse)) + 
  geom_tile() + 
  theme_bw() +
  scale_fill_gradient2(low = "blue",
    mid = "white", 
    high = "red", 
    midpoint =mean(res_db$test_rmse),
    space = "Lab", 
    na.value ="grey",
    guide = "colourbar", 
    aesthetics = "fill") + 
  labs(x = "Column Sample", y = "Subsample", fill = "RMSE")
g_8
```
A Column Sample of 0.6 and a Subsample of 0.7 minimize the test RMSE.

## XGBoost: Tuning the ETA Parameter
```{r}

set.seed(111111)
bst_mod_1 <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.03, 
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 1000, 
              early_stopping_rounds = 300, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)


set.seed(111111)
bst_mod_2 <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.01, 
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 1000, 
              early_stopping_rounds = 300, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)
set.seed(111111)
bst_mod_3 <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.005, 
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 1000, 
              early_stopping_rounds = 300, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)
set.seed(111111)
bst_mod_4 <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.001, 
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 1000, 
              early_stopping_rounds = 300, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)
set.seed(111111)
bst_mod_5 <- xgb.cv(data = dtrain, 
              
              nfold = 5, 
               
              eta = 0.0005, 
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 1000, 
              early_stopping_rounds = 300, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)


pd1 <- cbind.data.frame(bst_mod_1$evaluation_log[,c("iter", "test_rmse_mean")], rep(0.03, nrow(bst_mod_1$evaluation_log)))
names(pd1)[3] <- "eta"

pd2 <- cbind.data.frame(bst_mod_2$evaluation_log[,c("iter", "test_rmse_mean")], rep(0.01, nrow(bst_mod_2$evaluation_log)))
names(pd2)[3] <- "eta"

pd3 <- cbind.data.frame(bst_mod_3$evaluation_log[,c("iter", "test_rmse_mean")], rep(0.005, nrow(bst_mod_3$evaluation_log)))
names(pd3)[3] <- "eta"

pd4 <- cbind.data.frame(bst_mod_4$evaluation_log[,c("iter", "test_rmse_mean")], rep(0.001, nrow(bst_mod_4$evaluation_log)))
names(pd4)[3] <- "eta"

pd5 <- cbind.data.frame(bst_mod_5$evaluation_log[,c("iter", "test_rmse_mean")], rep(0.0005, nrow(bst_mod_5$evaluation_log)))
names(pd5)[3] <- "eta"

plot_data <- rbind.data.frame(pd1, pd2, pd3, pd4, pd5)

plot_data$eta <- as.factor(plot_data$eta)


g_9 <- ggplot(plot_data, aes(x = iter, y = test_rmse_mean, color = eta))+
  geom_smooth(alpha = 0.5) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        panel.border = element_blank(), 
        panel.background = element_blank()) + 
  labs(x = "Number of Trees", title = "Test RMSE v Number of Trees",
       y = "RMSE", color = "Learning \n Rate")
g_9

```
200 trees with an ETA of 0.01.

## XGBoost: Final Model
```{r}
set.seed(111111)
bst_3 <- xgboost(data = dtrain, 
              
              eta = 0.01,
              max.depth = 10, 
              min_child_weight = 10, 
              gamma = 0, 
              subsample = 0.7, 
              colsample_bytree =  0.6, 
               
              nrounds = 200, 
              early_stopping_rounds = 200, 
               
              verbose = 1,
              nthread = 6, 
              print_every_n = 20)

boost_preds <- predict(bst_3, dtest)

pred_dat <- cbind.data.frame(boost_preds, test_dat$point_diff)
names(pred_dat) <- c('prediction', 'actual')
pred_dat$error <- pred_dat$actual - pred_dat$prediction

pred_dat$winner <- 0
pred_dat$winner[pred_dat$prediction > 0] <- 1

rmse(boost_preds, test_dat$point_diff)

t <- table(pred_dat$winner, test_dat$winner)
confusionMatrix(t, positive = '1')



g_10 <- ggplot(pred_dat, aes(x = error)) +
   geom_density(alpha = 0.4, fill = 'blue') +
   theme(panel.background = element_blank(),
         panel.border = element_blank(),
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank()) +
   labs(title = 'Distribution of Error Terms',
        x = 'Value of Error Term',
        y = 'Frequency')
g_10

g_11 <- ggplot(pred_dat, aes(x = prediction, y = actual)) +
  geom_smooth() +
  geom_point() +
  theme(panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(title = 'Distribution of Error',
        x = 'Predicted Values',
        y = 'Actual Values')
g_11
```

